% !Mode::"TeX:UTF-8"
\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
%\usepackage{graphix}
\usepackage{listings}
\usepackage{booktabs}
\usepackage{amsthm}
\usepackage{CJK}
\usepackage{multirow}
    \newtheorem{theorem}{Theorem}[section]
\newcommand{\reffig}[1]{Fig.\ref{#1}}
%appendix
\usepackage[title,titletoc,toc]{appendix}
%format paragraphs
\usepackage{titlesec}
\titleformat{\paragraph}[runin]{\normalfont\normalsize\itshape}{\theparagraph}{}{}[.] %format paragraph italic and ser a period after it
\titlespacing{\paragraph}{0pt}{0pt}{*1} %remove spacing and add one characterspace after paragraph
\usepackage{subfigure}
\usepackage{epstopdf}
\usepackage{listings}
    \lstset{language=Matlab}%代码语言使用的是matlab
    \lstset{breaklines}%自动将长的代码行换行排版
    \lstset{extendedchars=false}%解决代码跨页时，章节标题，页眉等汉字不显示的问题


\author{Mubing Zhou}
%\date{2013-4-1}
\title{Matrix Derivatives with Chain Rule and Rules for Simple, Hadamard and Krnoecker Product}
\begin{document}
\maketitle

\section{Introduction}
Multivariable calculus has been a mature mathematical tool since 1970s, but people didn't have a very clear concept regarding to matrix derivative at that time. Though the chain rule and product rule for derivatives of composite scalar functions are very elementary result even for high school students, when it comes to matrix derivatives, people find it difficult. Simply transforming the results in scalar calculus to matrix calculus may yield wrong answers. For instance, the chain rule for a composite function is $\frac{\partial z}{\partial x} = \frac{\partial z}{\partial y} \frac{\partial y}{\partial x}$. But if we replace $x,y,z$ by matrices, the chain rule becomes invalid. The same situation is true for product rule. Moreover, not only simple product but other various products are defined, like Hadamard and Kronecker product. Finding the product rules for them is a even more hard work. So deriving rules for matrix calculus "would be virtually beyond the power of human thought to obtain in practice without the compact notation and the logical principles of the algebra of matrices". This paper, mainly based on the work of P.M. Bentler and S.Y. Lee, is devoted to seeking for the right expression for chain rule and the rules for simple, Hadamard and Kronecker product in matrix calculus.

In this paper, we will use the following notations:
\begin{itemize}
 % \item $I^n$: $n\times n$ identity matrix.
  \item $I^n$: $n\times n$ identity matrix.
  \item ${\overline X}: 1\times nm $ row vector of all elements of an $n\times m$ matrix $X$, taken row by row.
  \item $D_{{\overline X}}: nm\times nm$ diagonal matrix with diagonal elements $d_{ii} = x_i$, the corresponding element of the row vector ${\overline X}$.
  \item $X*Y$: Hadamard product of two $n\times m$ matrices $X$ and $Y$, i.e $X*Y = [x_{ij}y_{ij}]$.
  \item $X\otimes Y$: Kronecker product of two matrices, i.e., $X\otimes Y = [x_{ij}Y]$
  \item $E^{mn}$: $mn\times mn$ matrix such that, for $1\leq g \leq mn$ and $1\leq h\leq mn$, $e_{gh}=1$ if $g=n(j-1)+k,h=m(k-1)+j (0<j\leq m;0<k\leq n)$, and $e_{gh}=0$ otherwise.
\end{itemize}
Now we give the definition of matrix derivative. Let $Z$ be a $p\times q$ matrix and $Y$ be a $m\times n$ matrix whose elements are differentiable functions of all elements in $Z$. Define the matrix derivative of Y with respect to Z, $\frac{\partial Y}{\partial Z}$, to be a $pq\times mn$ matrix whose $i$th row is the mn-vector $\frac{\partial {\overline Y}}{\partial z_i}$, where $z_i$ is the i-th element in ${\overline Z}$. Obviously from the definition, $\frac{\partial Y}{\partial Z} = \frac{\partial {\overline Y}}{\partial {\overline Z}}$.

By now, it is necessary to give some examples to illustrate the above concepts. Suppose $Z = (z_1, z_2), Y = \left( {\begin{array}{*{20}{c}}
  y_1&y_2 \\
  y_3&y_4
\end{array}} \right)$ and $X = \left( {\begin{array}{*{20}{c}}
  x_1&x_2 \\
  x_3&x_4
\end{array}} \right)$. Then
\begin{eqnarray*}
% \nonumber to remove numbering (before each equation)
  {\overline Y} &=& (y_1, y_2, y_3, y_4) \\
  \\
  D_{{\overline Y}}&=& diag\{y_1, y_2, y_3, y_4\} \\
  \\
  X*Y &=& \left( {\begin{array}{*{20}{c}}
                x_1y_1&x_2y_2 \\
                x_3y_3&x_4y_4
            \end{array}} \right) \\
            \\
  Z\otimes Y &=& (z_1 Y, z_2Y) \\
             &=& \left( {\begin{array}{*{20}{c}}
                z_1y_1&z_1y_2  & z_2y_1&z_2y_2 \\
                z_1y_3&z_1y_4  & z_3y_3&z_2y_4
                \end{array}} \right) \\
\end{eqnarray*}
, where $*$ and $\otimes$ denote Hadamard product and Krnoecker product respectively. Then the derivative of $Y$ with respect to $Z$ can be shown to be
\begin{eqnarray*}
% \nonumber to remove numbering (before each equation)
    \frac{\partial Y}{\partial Z} &=&
   \left( {\begin{array}{*{20}{c}}
     \frac{\partial y_1}{\partial z_1}& \frac{\partial y_2}{\partial z_1}& \frac{\partial y_3}{\partial z_1}& \frac{\partial y_4}{\partial z_1} \\\\
     \frac{\partial y_1}{\partial z_2}& \frac{\partial y_2}{\partial z_2}& \frac{\partial y_3}{\partial z_2}& \frac{\partial y_4}{\partial z_2} \\
    \end{array}} \right) \\
\end{eqnarray*}

\section{Chain rule and simple product rule}
The chain rule is given in the following theorem. 
\begin{theorem}\label{chainrule_thm}
Suppose each element of a \textbf{vector} $Y=[Y_1, Y_2]$ is a differential function of all elements of a \textbf{vector} $Z$, where $Y_1$ contains all the mathematical variables which are independent to each other in $Y$ and $Y_2$ are functions of elements in $Y_1$. Furthermore, suppose that each element of a \textbf{vector} $X$ is a differentiable function of all elements of $Y_1$. Then
\begin{equation}\label{chainrule}
\frac{\partial X}{\partial Z} = \frac{\partial Y_1}{\partial Z} \frac{\partial X}{\partial Y_1}
\end{equation}
\end{theorem}

\begin{proof}
Suppose $X$ is $1\times r$, $Y_1$ is $1\times s$ and $Z$ is $1\times t$. Then we have
\[\frac{\partial X}{\partial Z} =
\left( {\begin{array}{*{20}{c}}
\frac{\partial x_1}{\partial z_1} & \cdots & \frac{\partial x_r}{\partial z_1}  \\
\vdots & \ddots& \vdots \\
\frac{\partial x_1}{\partial z_t}& \cdots & \frac{\partial x_r}{\partial z_t}
\end{array}} \right)\]
and
\[\frac{\partial Y_1}{\partial Z} \frac{\partial X}{\partial Y_1} =
\left( {\begin{array}{*{20}{c}}
\frac{\partial y_1}{\partial z_1} & \cdots & \frac{\partial y_s}{\partial z_1}  \\
\vdots & \ddots& \vdots \\
\frac{\partial x_1}{\partial z_t}& \cdots & \frac{\partial y_s}{\partial z_t}
\end{array}} \right)
\left( {\begin{array}{*{20}{c}}
\frac{\partial x_1}{\partial y_1} & \cdots & \frac{\partial x_r}{\partial y_1}  \\
\vdots & \ddots& \vdots \\
\frac{\partial x_1}{\partial y_s}& \cdots & \frac{\partial x_r}{\partial y_s}
\end{array}} \right)
\]
The (i,j)th element in $\frac{\partial X}{\partial Z}$ is $\frac{\partial x_i}{\partial z_j}$ while the corresponding element in $\frac{\partial Y_1}{\partial Z} \frac{\partial X}{\partial Y_1}$ is $\sum \limits_{l=1}^s \frac{\partial y_l}{\partial z_i} \frac{\partial x_j}{\partial y_l}$. By the chain rule in scalar calculus, we verify Eq.\ref{chainrule}.
\end{proof}

Note that in Theorem.\ref{chainrule_thm}, we split $Y$ into two parts. If we don't do that, the result may be wrong. Generally speaking, the following equation is NOT true:
\[\frac{\partial X}{\partial Z} = \frac{\partial Y}{\partial Z} \frac{\partial X}{\partial Y} \]
A simple counterexample is to take $X=[v_1, v_1, v_2, v_2],$ $Y=[v_1, v_2, v_1, v_2]$ and $Z = [v_1, v_2, v_2, v_1]$. Now
$ \frac{\partial X}{\partial Z} =
\left( {\begin{array}{*{20}{c}}
1& 1& 0& 0 \\
0& 0& 1& 1 \\
0& 0& 1& 1 \\
1& 1& 0& 0
\end{array}} \right)
$
But $\frac{\partial Y}{\partial Z} \frac{\partial X}{\partial Y} = \left( {\begin{array}{*{20}{c}}
2& 2& 0& 0 \\
0& 0& 2& 2 \\
0& 0& 2& 2 \\
2& 2& 0& 0
\end{array}} \right)$
. Note that here $Y_1 = [v_1, v_2]$.

The product rule is obtain in Theorem.\ref{productrule_thm}
\begin{theorem}\label{productrule_thm}
Suppose each element of a $p\times r$ matrix $Y$ and an $r\times q$ matrix $X$ is a differentiable function of all elements of a $m\times n$ matrix $Z$. Then
\begin{equation}\label{productrule}
  \frac{\partial(YX)}{\partial Z} = \frac{\partial Y}{\partial Z} (I^p \otimes X) + \frac{\partial X}{\partial Z}(Y' \otimes I^q)
\end{equation}
\end{theorem}

\begin{proof}
$\frac{\partial(YX)}{\partial Z}$ is a $mn\times pq$ matrix. Its (i,j)th element corresponds to the i-th element in ${\overline Z}$, say $z_\beta$, and j-th element in $\overline {XY}$, say $(XY)_{(s,t)}$. $(XY)_{(s,t)}$ is the product of the s-th row of $Y$ and t-th column of $X$, or $(XY)_{(s,t)} = \sum \limits_{l=1}^r y_{sl}x_{lt}$. So
\[ \frac{\partial(YX)}{\partial Z} = [\frac{\partial \sum \limits_{l=1}^r y_{sl}x_{lt}}{\partial z_\beta}] = \sum\limits_{l=1}^r \frac{\partial y_{sl}}{\partial z_\beta}x_{lt} + \frac{\partial x_{lt}}{\partial z_\beta}y_{sl}\]

Now consider $\frac{\partial Y}{\partial Z}(I^p \otimes X)$. $\frac{\partial Y}{\partial Z}$ is a $mn\times pr$ matrix while $I^p \otimes X$ is a $pr\times pq$ matrix. The (i,j)th element of $\frac{\partial Y}{\partial Z}(I^p \otimes X)$ is the product of i-th row of $\frac{\partial Y}{\partial Z}$ and j-th column of $I^p \otimes X$. The i-th row of corresponds to the i-th element of $\overline {Z}$, which is $z_\beta$. Recall that $I^p \otimes X$ is in the form of
$ \left( {\begin{array}{*{20}{c}}
X & & & \\
 &X & & \\
 &  &\ddots& \\
 & & & X
\end{array}} \right)
$
So the j-th column of $I^p \otimes X$ corresponds to the t-column of $X$. And the t-th column of $X$ should be multiplied by the s-th row of $Y$. So the (i,j)th element of $\frac{\partial Y}{\partial Z}(I^p \otimes X)$ is $\sum \limits_{l=1}^k \frac{\partial y_{sl}}{\partial z_\beta} x_{lt}$. Similarly, we can prove that the (i,j)th element of $\frac{\partial X}{\partial Z}(Y' \otimes I^q)$ is $\sum \limits_{l=1}^k \frac{\partial x_{lt}}{\partial z_\beta}y_{sl}$. Add these two parts up is just the (i,j)th element of $\frac{\partial(YX)}{\partial Z}$.
\end{proof}



 \end{document}

